---
layout: post
title: Basic regression in Gaussian processes 
category: ecology
tag: nonparametric-bayes
code: true
output: html_document
---

First I take a look at doing some simple tasks involving Gaussian processes in base R and sample from them with basic multi-variate normal random draw, `mvrnorm`.  Then, I speculate (unsuccessfully) on how to do this in greta.


```{r setup, message = FALSE, warning = FALSE}

require(MASS)
require(reshape2)
require(ggplot2)

knitr::opts_chunk$set(message=FALSE, warning=FALSE)


set.seed(12345)


```



```{r}
x_predict <- seq(-5,5,len=50)
l <- 1
```

We will use the squared exponential (also called radial basis or Gaussian, though it is not this that gives Gaussian process it's name; any covariance function would do) as the covariance function, 

$$\operatorname{cov}(X_i, X_j) = \exp\left(-\frac{(X_i - X_j)^2}{2 \ell ^ 2}\right)$$


```{r}
SE <- function(Xi,Xj, l) exp(-0.5 * (Xi - Xj) ^ 2 / l ^ 2)
cov <- function(X, Y) outer(X, Y, SE, l)
```

```{r}
COV <- cov(x_predict, x_predict)
```

Generate (draw) a number of functions from the process (note that this is not conditioned on any data, and we are leaving the hyperparameter `l` fixed)


```{r}
values <- mvrnorm(200, rep(0, length=length(x_predict)), COV)
```


Reshape the data into long (tidy) form, listing x value, y value, and sample number

```{r}
dat <- data.frame(x=x_predict, t(values)) %>%
  tidyr::pivot_longer(-x, names_to = "rep", values_to = "value") %>% 
  mutate(rep = as.numeric(as.factor(rep)))

ggplot(dat,aes(x=x,y=value)) +
  geom_line(aes(group=rep), color =  rgb(0.7, 0.1, 0.4), alpha = 0.4) 
```



### Posterior distribution conditional on data


Imagine we have some data,


```{r}
obs <- data.frame(x = c(-4, -3, -1,  0,  2),
                  y = c(-2,  0,  1,  2, -1))
```



In general we aren't interested in drawing from the prior, but want to include information from the data as well.  We want the joint distribution of the observed values and the prior is:

$$\begin{pmatrix} y_{\textrm{obs}} \\ y_{\textrm{pred}} \end{pmatrix} \sim \mathcal{N}\left( \mathbf{0}, \begin{bmatrix} cov(X_o,X_o) & cov(X_o, X_p) \\ cov(X_p,X_o) & cov(X_p, X_p) \end{bmatrix} \right)$$


No observation noise
--------------------

Assuming the data are known without error and conditioning on the data, and given $x \sim \mathcal{N}(0, cov(X_o, X_o))$, then the conditional probability of observing our data is easily solved by exploiting the nice properties of Gaussians,

$$x|y \sim \mathcal{N}\left(E,C\right)$$
$$E = cov(X_p, X_o) cov(X_o,X_o)^{-1} y$$
$$C= cov(X_p, X_p) - cov(X_p, X_o) cov(X_o,X_o)^{-1} cov(X_o, X_p) $$

(We use `solve(M)` which with no second argument will simply invert the matrix `M`, but should use the Cholsky decomposition instead for better numerical stability)


```{r}
cov_xx_inv <- solve(cov(obs$x, obs$x))
Ef <- cov(x_predict, obs$x) %*% cov_xx_inv %*% obs$y
Cf <- cov(x_predict, x_predict) - cov(x_predict, obs$x)  %*% cov_xx_inv %*% cov(obs$x, x_predict)
```



Now lets take 3 random samples from the posterior distribution,

```{r}
values <- mvrnorm(200, Ef, Cf)
```


and plot our solution (mean, 2 standard deviations, and the ranom samples.)


```{r}
dat <- data.frame(x=x_predict, t(values)) %>%
  tidyr::pivot_longer(-x, names_to = "rep", values_to = "value") %>% 
  mutate(rep = as.numeric(as.factor(rep)))


gp <- data.frame(x = x_predict, Ef = Ef, sigma = 2*sqrt(diag(Cf)) )

ggplot(dat,aes(x=x,y=value)) + 
  #geom_line(aes(group=rep), color =  rgb(0.7, 0.1, 0.4), alpha = 0.2) + #REPLICATES
  geom_ribbon(data = gp, 
              aes(x, 
                  y = Ef, 
                  ymin = Ef - sigma, 
                  ymax = Ef + sigma),
              fill="blue", alpha = 0.1) +
  geom_line(dat = gp, aes(x=x,y=Ef), size=1) + #MEAN
  geom_point(data=obs,aes(x=x,y=y)) +  #OBSERVED DATA
  scale_y_continuous(lim=c(-3,3), name="output, f(x)") +
  xlab("input, x")
```






Additive noise
--------------

In general the model may have process error, and rather than observe the deterministic map $f(x)$ we only observe $y = f(x) + \varepsilon$.  Let us assume for the moment that $\varepsilon$ are independent, normally distributed random variables with variance $\sigma_n^2$.  


```{r}
sigma.n <- 0.25
cov_xx_inv <- solve(cov(obs$x, obs$x) + sigma.n^2 * diag(1, length(obs$x)))
Ef <- cov(x_predict, obs$x) %*% cov_xx_inv %*% obs$y
Cf <- cov(x_predict, x_predict) - cov(x_predict, obs$x)  %*% cov_xx_inv %*% cov(obs$x, x_predict)
```




Now lets take 3 random samples from the posterior distribution,

```{r}
values <- mvrnorm(200, Ef, Cf)
```


and plot 


```{r}
dat <- data.frame(x=x_predict, t(values))
dat <- melt(dat, id="x")

gp <- data.frame(x_predict = x_predict, Ef = Ef, sigma = 2*sqrt(diag(Cf)) )


ggplot(dat,aes(x=x,y=value)) +
  geom_ribbon(data=gp, 
              aes(x=x_predict, y=Ef, ymin=Ef-sigma, ymax=Ef+sigma),
              fill="grey80") + # Var
  geom_line(aes(color=variable)) + #REPLICATES
  geom_line(data=gp,aes(x=x_predict,y=Ef), size=1) + #MEAN
  geom_point(data=obs,aes(x=x,y=y)) +  #OBSERVED DATA
  scale_y_continuous(lim=c(-3,3), name="output, f(x)") +
  xlab("input, x")
```





------

# greta

Can we express the same operations in greta?  

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(greta)
library(greta.gp)
```




```{r}
# kernel & GP
y <- obs$y
x <- obs$x

# hyperparameters
rbf_var <- 10 # lognormal(0, 1)
rbf_len <- 0.5 #lognormal(0, .01)
obs_sd <- .01 #lognormal(0, .01)

# kernel & GP
kernel <- rbf(rbf_len, rbf_var)
f <- gp(x, kernel)

# likelihood -- not clear how this influences f_plot?
distribution(y) <- normal(f, obs_sd)

# prediction
f_plot <- project(f, x_plot)
```


```{r}
# fit the model by Hamiltonian Monte Carlo
m <- model(f_plot)
draws <- mcmc(m, warmup = 0)
```
```


